---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(343)
```

```{r setup}
library(fabricatr)
```

fabricatr makes data you do not have yet. You describe the variables and how they relate, and `fabricate()` builds a data frame that looks like what you expect to collect, so a design can be run and diagnosed before there is anything to analyze.

## One level

`fabricate()` builds a data frame one column at a time. `N` is the number of rows, and every column can use the columns before it:

```{r}
survey <- fabricate(
  N = 100,
  age = round(runif(N, 18, 80)),
  income = 20 + 0.5 * age + rnorm(N, sd = 10)
)
head(survey)
```

An `ID` column comes for free. `N` is available inside every expression, which is why `rnorm(N)` draws one value per row.

## Two levels

Real data have structure: citizens in villages, students in schools, years within countries. `add_level()` builds the top level and `nest_level()` puts children inside each parent. A column at the village level is copied down to every citizen in the village, so a citizen's income can depend on where they live:

```{r}
nested <- fabricate(
  villages = add_level(N = 5, village_wealth = rnorm(N)),
  citizens = nest_level(N = 4, income = village_wealth + rnorm(N))
)
head(nested)
```

Inside `citizens`, `N` is 20, the number of citizens being built, not 4. Each level names its own ID column after itself.

## Potential outcomes

An experiment has an outcome under treatment and an outcome under control, and observes only one of them. `potential_outcomes()` writes both down from a formula, one column per condition:

```{r}
experiment <- fabricate(
  N = 100,
  U = rnorm(N),
  potential_outcomes(Y ~ 0.5 * Z + U)
)
head(experiment)
```

Once units are assigned, `reveal_outcomes()` picks out the outcome each unit shows:

```{r}
observed <- fabricate(
  data = experiment,
  Z = rbinom(N, 1, 0.5),
  Y = reveal_outcomes(Y ~ Z)
)
head(observed)
```

## Draws that look like real variables

The `draw_*` family turns a latent variable into the kind of measurement a survey produces. `draw_binary()` gives a 0/1 outcome, here through a logit link:

```{r}
fabricate(
  N = 6,
  x = rnorm(N),
  voted = draw_binary(latent = x, link = "logit")
)
```

`draw_normal_icc()` draws a clustered variable with a target intra-cluster correlation, so that citizens in the same village resemble each other by the amount you choose:

```{r}
clustered <- fabricate(
  villages = add_level(N = 20),
  citizens = nest_level(N = 10, trust = draw_normal_icc(clusters = villages, ICC = 0.3))
)
summary(lm(trust ~ villages, data = clustered))$r.squared
```

## Resampling

`resample_data()` bootstraps a data frame, and respects its levels: sample villages, then citizens within the villages drawn.

```{r}
boot <- resample_data(nested, N = c(villages = 3, citizens = 2))
boot
```

`ALL` in place of a number keeps every unit at that level.

## Where to read next

The book at <https://book.declaredesign.org> puts fabricated data to work in complete research designs, and `vignette("fabricatr2.0")` lists what changed from fabricatr 1.x for anyone with an older script.
