---
title: "Using other data generating packages with **fabricatr**"
author: "Aaron Rudkin"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using other data generating packages with **fabricatr**}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
options(digits=2)
set.seed(19861108)
library(fabricatr)
```

In general, **fabricatr** is going to be compatible with any existing packages you use to generate synthetic data in one of two ways: either using those packages to create variables within a `fabricate` call, or using those packages to make complete data frames which are then imported into a `fabricate`. Below we provide examples for some of the most popular packages that serve this purpose. Please feel free to [http://discuss.declaredesign.org/](Contact Us) if you have any questions about using another package in concert with **fabricatr**.

## **wakefield**: simulating common demographic features

**wakefield** (by [Tyler Rinker](https://github.com/trinker)) is a popular `R` package for creating synthetic data. **wakefield**'s strength is that it can quickly generate common variables, especially for human demographic features. **wakefield** can easily be integrated into a **fabricatr** workflow in one of two ways: using **wakefield** to create individual variables within a `fabricate` call, or using **wakefield** to make a data frame and importing that data frame into a `fabricate` call.

In this example, we create a data-set of participants in a survey experiment, using **wakefield** to generate the demographic variables

```{r eval=FALSE}
library(wakefield)

survey_experiment_df <- fabricate(
  N = 50,
  treatment = draw_binary(prob = 0.5, N = N),
  age = age(n = N),
  race = race(n = N),
  sex = sex(n = N),
)
```

Researchers interested in learning about **wakefield**'s available functionality, parameterizations, and default probability can [read **wakefield**'s user guide on GitHub](https://github.com/trinker/wakefield).

In addition to creating variables within a `fabricate` call, users can import completed **wakefield** data frames into a `fabricate` call:

```{r eval=FALSE}
survey_experiment_df <- r_data_frame(
  n = 50,
  age,
  race,
  sex)

fabricatr_df <- fabricate(
  data = survey_experiment_df,
  treatment = draw_binary(prob = 0.5, N = N)
)
```

## **randomNames**: Plausible names for human subjects

**randomNames** (by [Damian Betebenner](https://github.com/dbetebenner)) is a package that does one thing well: generate random names for human subjects (including with specified genders and ethnicities). The primary use case would be to use this as part of generating a variable within a `fabricate` call. In the below example, we use `fabricate` to generate some other demographic data, and then `randomNames` to generate matching names.

```{r eval=FALSE}
library(randomNames)

experiment_data <- fabricate(
  N = 50,
  treatment = draw_binary(prob = 0.5, N = N),
  is_female = draw_binary(prob = 0.5, N = N),
  patient_name = randomNames(N, gender=is_female)
)
```

Note that we make use of the existing `is_female` variable from the `fabricate` call to ensure `randomNames` generates gender-typical names.

## Modeling causality with DAGs and **simcausal**

Users who are familiar with the DAGS (directed acyclic graphs) model of causal inference may have interest in using the **simcausal** package, which allows users to specify a DAGS model and then sample from it. Integrating this package with **fabricatr** is likely to involve using **simcausal** first to specify a model, simulating data from the model, and then importing data into a `fabricate` call for further user with **fabricatr**.

Consider this example, common in the literature on educational attainment and school outcomes, where students come from families that have a `wealth` parameter, assignment to schools is based partially on `wealth`, and test outcomes (`testoutcome`) is based on both school quality and wealth.

```{r eval=FALSE}
library("simcausal")

# Define DAG
D <- DAG.empty() + 
  node("wealth", distr = "rnorm",
       mean = 30000,
       sd = 10000) +
  node("schoolquality", distr = "runif",
       min = 0 + (5 * (wealth > 50000)),
       max = 10) +
  node("testoutcome", distr = "runif",
       min = 0 + 0.0001 * wealth + 0.25 * schoolquality,
       max = 10)

# Freeze DAG object
set_dag <- set.DAG(D)

# Draw data from DAG
df <- sim(set_dag, n = 100)

# Pass into fabricate call and make new variables as necessary
fabricate(df,
          passed_test = testoutcome > 6,
          eligible_for_snap = wealth < 25000)
```

## Survival and duration models with **simsurv**

**simsurv** is a package dedicated to generating panel survival data. The most likely way you might integrate **simsurv** with **fabricatr** would be to use **fabricatr** to generating covariates which can then be imported into **simsurv** to model in a hazard or duration context.

Here, our example will be a clinical trial of a cancer drug. Participants have the expected biographical data: age, gender, whether the patient smokes, the disease stage, assignment to treatment, and a KPS score (commonly used to evaluate overall patient health).

Survival data creates a ragged longitudinal survey; some patients will die during the course of the trial, removing them as observations. Others will continue alive until the end of the trial. We specify a "hazard function", which tells **simsurv** how the course of patient survival will change over time. Covariates with positive `betas` increase risk of death, while covariates with negative `betas` decrease risk of death. We will track patients for 5 years after treatment.

```{r eval=FALSE}
library(simsurv)

# Simulate patient data in a clinical trial
participant_data <- fabricate(
  N = 100,
  age = runif(N, min = 18, max = 85),
  is_female = draw_binary(prob = 0.5, N = N),
  is_smoker = draw_binary(prob = 0.2 + 0.2 * (age > 50), N = N),
  disease_stage = round(runif(N, min = 1 + 0.5 * (age > 65), max = 4)),
  treatment = draw_binary(prob = 0.5, N = N),
  kps = runif(N, min = 40, max = 100)
)

# Simulate data in the survival context
survival_data <- simsurv(
  lambdas = 0.1, gammas = 0.5,
  x = participant_data, 
  betas = c(is_female = -0.2, is_smoker = 1.2,
            treatment = -0.4, kps = -0.005,
            disease_stage = 0.2),
  maxt = 5)
```

The generated data from the `survival_data` object can then be re-imported into the `participant_data` using any data merging tools, including through a `fabricate` call, and then used for subsequent analyses (e.g. using the **survival** package).

## Time series using **forecast**

**forecast**, by [Rob Hyndman](https://github.com/robjhyndman/forecast), is a package commonly used to analyze time series data which also has functionality capable of generating simulated time series data. **forecast** can use the `Arima` and `simulate` functions to create pre-specified ARIMA models, including seasonal time trends.

Below, we provide an example of using **forecast** to generate an ARIMA time series, reshape the data, and import it into **fabricatr** to create new variables of interest.

```{r eval=FALSE}
library(forecast)

arima_model <- simulate(
  Arima(ts(rnorm(100), frequency = 4),
        order = c(1, 0, 1))
  
fabricate(data.frame(arima_model), 
          year = rep(1:25, each=4),
          quarter = rep(1:4, 25))
```

Here, `ts` converts a series of data into a time series, with frequency specifying the number of observations per unit of time (in this case, for example, quarters in a year). `Arima` ingests this data and fits an ARIMA model with the specified parameters. `simulate` draws new data from the fit time series, producing a vector of interest. We then import the data into a `fabricate` call (converting it to a data frame) and add new columns of interest.

## Other data simulation tools

The `R` ecosystem has many other data simulation tools, and all can be used to complement or supplement **fabricatr** in your workflow. Some of the packages that we have noticed but not covered here include:

* [gems](https://cran.r-project.org/web/packages/gems/index.html) by Luisa Salazar Vizcaya
* [simFrame](https://cran.r-project.org/web/packages/simFrame/index.html) by Andreas Alfons
* [simPop](https://cran.r-project.org/web/packages/simPop/index.html) by Matthias Templ
* [simStudy](https://cran.r-project.org/web/packages/simstudy/index.html) by Keith Goldfeld
* [synthPop](https://cran.r-project.org/web/packages/synthpop/index.html) by Beata Nowok

If you'd like to see a tutorial on using these packages or any others with **fabricatr**, please [Contact Us](http://discuss.declaredesign.org/) so we can help you
