---
title: "Resampling data with fabricatr"
author: "Aaron Rudkin"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Resampling data with fabricatr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
options(digits=2)
set.seed(19861108)
library(fabricatr)
```

One way to imagine new data is to take data you already have and resample it, ensuring that existing inter-correlations between variables are preserved, while generating new data or expanding the size of the dataset. **fabricatr** offers several options to simulate resampling.

# Bootstrapping

The simplest option in **fabricatr** is to "bootstrap" data. Taking data with N observations, the "bootstrap" resamples these observations with replacement and generates N new observations. Existing observations may be used zero times, once, or more than once. Bootstrapping is very simple with the `resample_data()` function:

```{r echo=TRUE, results="hide"}
survey_data <- fabricate(N = 10, voted_republican = draw_binary(prob = 0.5, N = N))

survey_data_new <- resample_data(survey_data)
head(survey_data_new)
```
```{r echo=FALSE}
knitr::kable(head(survey_data_new))
```

It is also possible to resample fewer or greater number of observations from your existing data. We can do this by specifying the argument `N` to `resample_data()`. Consider expanding a small dataset to allow for better imagination of larger data to be collected later.

```{r echo=TRUE, results="hide"}
large_survey_data <- resample_data(survey_data, N = 100)
nrow(large_survey_data)
```
`r nrow(large_survey_data)`

# Resampling hierarchical data

One of the most powerful features of all of **fabricatr** is the ability to resample from hierarchical data at any or all levels. Doing so requires specifying which levels you will want to resample with the `ID_labels` argument. Unless otherwise specified, all units from levels below the resampled level will be kept. In our earlier country-province-citizen dataset, resampling countries will lead to all provinces and citizens of the selected country being carried forward. You can resample at multiple levels simultaneously.

Consider this example, which takes a dataset containing 2 cities of 3 citizens, and resamples it into a dataset of 3 cities, each containing 5 citizens.

```{r echo=TRUE, results="hide"}
my_data <-
  fabricate(
    cities = add_level(N = 2, elevation = runif(n = N, min = 1000, max = 2000)),
    citizens = add_level(N = 3, age = runif(N, 18, 70))
  )

my_data_2 <- resample_data(my_data, N = c(3, 5), ID_labels = c("cities", "citizens"))
head(my_data_2)
```
```{r echo=FALSE}
knitr::kable(head(my_data_2))
```

`resample_data()` will first select the cities to be resampled. Then, for each city, it will continue by selecting the citizens to be resampled. If a higher level unit is used more than once (for example, the same city being chosen twice), and a lower level is subsequently resampled, the choices of which units to keep for the lower level will differ for each copy of the higher level. In this example, if city 1 is chosen twice, then the sets of five citizens chosen for each copy of the city 1 will differ.

You can also specify the levels you wish to resample from using the name arguents to the `N` parameter, like in this example which does exactly the same thing as the previous example, but specifies the level names in a different way:

```{r echo=TRUE, results="hide"}
my_data <-
  fabricate(
    cities = add_level(N = 2, elevation = runif(n = N, min = 1000, max = 2000)),
    citizens = add_level(N = 3, age = runif(N, 18, 70))
  )

my_data_2 <- resample_data(my_data, N = c(cities = 3, citizens = 5))
head(my_data_2)
```
```{r echo=FALSE}
knitr::kable(head(my_data_2))
```

# "Passthrough" Resampling

In some cases it may make sense to resample each unit at a given level. For example, there may be value in resampling 1 citizen in each and every city represented in the data set. **fabricatr** allows the user to specify `ALL` for the `N` argument to a given level to accomplish this:

```{r echo=TRUE, results="hide"}
my_data <-
  fabricate(
    cities = add_level(N = 2, elevation = runif(n = N, min = 1000, max = 2000)),
    citizens = add_level(N = 3, age = runif(N, 18, 70))
  )

my_data_3 <- resample_data(my_data, N = c(ALL, 1), ID_labels = c("cities", "citizens"))
head(my_data_3)
```
```{r echo=FALSE}
knitr::kable(head(my_data_3))
```
