---
title: "Getting Started with fabricatr: Resampling Data"
author: "Aaron Rudkin"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with Fabricatr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(fabricatr)
```

`fabricatr` is a package designed to help you imagine your data before you collect it. While many solutions exist for creating simulated datasets, `fabricatr` is specifically designed to make the creation of realistic social science datasets easy. In particular, we need to be able to imagine **correlated** data and **hierarchical** data. `fabricatr` is designed to integrate into a [tidyverse](https://www.tidyverse.org/) workflow, and to allow users to imagine data from scratch or by modifying existing data.

`fabricatr` is a member of the `DeclareDesign` software suite that also includes the `r` packages [randomizr](randomizr.declaredesign.org), [estimatr](estimatr.declaredesign.org), and [Declare Design](declaredesign.org). 

# Simulating "resampling" from existing data.

One way to imagine new data is to take data you already have and resample it, ensuring that existing inter-correlations between variables are preserved, while generating new data or expanding the size of the dataset. `fabricatr` offers several options to simulate resampling.

# Bootstrapping

The simplest option in `fabricatr` is to "bootstrap" data. Taking data with N observations, the "bootstrap" resamples these observations with replacement and generates N new observations. Existing observations may be used zero times, once, or more than once. Bootstrapping is very simple with the `resample_data` function:

```{r}
survey_data = fabricate(N=10, 
                        voted_republican = draw_binary(N=N, x=0.5))

survey_data_new = resample_data(survey_data)
survey_data
```

It is also possible to resample fewer or greater number of observations from your existing data. We can do this by specifying the argument `N` to `resample_data`. Consider expanding a small dataset to allow for better imagination of larger data to be collected later.

```{r}
large_survey_data = resample_data(survey_data, N=100)
nrow(large_survey_data)
```

# Resampling hierarchical data

One of the most powerful features of all of `fabricatr` is the ability to resample from hierarchical data at any or all levels. Doing so requires specifying which levels you will want to resample with the `ID_labels` argument. Unless otherwise specified, all units from levels below the resampled level will be kept. In our earlier country-province-citizen dataset, resampling countries will lead to all provinces and citizens of the selected country being carried forward. You can resample at multiple levels simultaneously.

Consider this example, which takes a dataset containing 2 cities of 3 citizens, and resamples it into a dataset of 3 cities, each containing 5 citizens.

```{r}
my_data <-
  fabricate(
    cities = level(N = 2, elevation = runif(n = N, min = 1000, max = 2000)),
    citizens = level(N = 3, age = runif(N, 18, 70))
  )

my_data_2 <- resample_data(my_data, 
                           N = c(3, 5), 
                           ID_labels = c("cities", "citizens"))
my_data_2
```

`resample_data` will first select the cities to be resampled. Then, for each city, it will continue by selecting the citizens to be resampled. If a higher level unit is used more than once (for example, the same city being chosen twice), and a lower level is subsequently resampled, the choices of which units to keep for the lower level will differ for each copy of the higher level. In this example, if city 1 is chosen twice, then the sets of five citizens chosen for each copy of the city 1 will differ.
